### 2019-11-27 paper/conceptual_framework.tex
Edit and shorten conceptual framework
Below 2,000 words now. Still seems too long.

This section provides a conceptual framework for understanding what determines the  spacing 
between births.
I first discuss the main determinants of spacing discussed in the prior research, and 
then turn to how son preference and sex selection fits into the decisions on spacing.
Since sex selection and education are closely linked, I pay special attention to how education affects spacing.

In the simplest possible dynamic model of fertility, the spacing decision is trivial.
Assume that parents derive benefits from the presence of children, rather than from 
the births themselves, that spacing of children by itself does impact the parental utility, 
and that the parents perfectly control when to have children. 
In this model, the optimal decision is then to have the children as early as possible 
\citep{Newman1988}.
The simplest model does not fit the observed spacing patterns, which suggests 
that one or more of the assumptions in the simple model are wrong.

First, parents do not entirely control the timing of births, but rather decide on the
levels of sexual activity and contraceptive effort, which, in turn, affect the 
likelihood of conceiving.

Couples make decisions on sexual activity and contraception, taking into account fecundity, determined by factors such as age and health of the mother, and their knowledge of and access to contraceptive technology.
There is evidence from ``natural fertility'' populations that higher fertility is associated with shorter spacing and that birth intervals increase 
with parity \citep{Henry1961,Leridon1977}.

Age and health of the mother are, however, unlikely to be significant factors in changes
in spacing over time or across education levels in India for the period covered, since
there is little evidence that women's health and nutritional status substantially affect 
spacing, outside of direct famine conditions \citep{Huffman1987,John1987,lindstrom99}.
The Maharashtra drought occurred in 1970--1973 and would, therefore, only impact the first
year or two of the data.
Furthermore, I only examine spacing up to parity fourth, and the age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, even though better -educated women are, on average, older when they marry.

Despite a general belief that better access to and knowledge of contraception should lead
to longer average spacing, the empirical relationship between contraception use and birth 
spacing is ambiguous \citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
This lack of evidence is consistent with the ambiguous theoretical effect of introducing 
modern contraceptives on birth spacing.
On the one hand, in the absence of modern contraceptive technologies, parents may try to 
avoid overshooting their desired number of children by spacing their children further apart.
Increased reliability of the access to and effectiveness of contraceptives then lead to 
shorter intervals between births because parents are more confident of effectively
stopping childbearing when they reach their preferred number of children and having
the children closer together provide higher utility because parents benefit from the
presence of children for longer as discussed above \citep{Keyfitz1971,Heckman1976}.
On the other hand, when contraceptives are effective, we would expect some birth 
intervals to be shorter than ideal, as discussed below, through simple random
variation.
With increased reliability of the  access to and effectiveness of contraceptives, parents 
would better be able to avoid too short spacing between birth and average spacing
should go up.

To further complicate the relationship between contraception and spacing, the ability to 
successfully use ``low efficacy'' contraceptive methods, such as the rhythm method, is 
increasing in education, while there is no difference across education levels 
when it comes to modern contraceptives such as pills or injectables \citep{Rosenzweig1989}.
Furthermore, even in the absence of modern contraceptives, families appear able to time or 
at least postpone births when they have substantial enough incentive 
\citep{Jayachandran2011,Alam2018}.

Second, parents may care about the spacing between birth because of the potential effects of spacing on the children.%
\footnote{
In principle, the mother could also be affected, but the results
so are mixed on mothers' 
anthropometric status, and there appears to be no effect on mortality 
\citep{Ronsmans1998,Menken2003,Dewey2007,Conde-Agudelo2012}.
}
One theory is that child ``quality'' might increases with spacing 
because the amount of parental attention increases with spacing
\citep{Zajonc1975,Zajonc1976,Razin1980}.
In this case, birth spacing should increase in family income, while the effect of maternal 
education is ambiguous since higher education, on the one hand, makes the mother's time spent 
out of the labor market more costly and on the other hand increases how productive the 
mother is in childrearing. 
For child quality measures such as IQ and education, there is, to my knowledge, no evidence 
from developing countries.%
\footnote{
The evidence from developed countries is mixed
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.
}

For one crucial aspect of child quality---health and mortality---shorter spacing 
does lead to worse outcomes, especially for very short intervals of 24 months or less,
although this effect decreases with maternal education
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
It is also possible the underlying mortality risk affects birth 
spacing.
The predicted outcome then---ignoring the effect of shorter spacing on mortality risk---would be to shorten 
birth spacing to get to the ideal number of children earlier.
There is little empirical evidence for this effect \citep{Newman1983,Newman1988,Bhalotra2008}.
}
Hence, if parents realize short birth spacing increases their children's mortality risk, 
they likely prefer to space births further apart.
However, since women with secondary or tertiary education show relatively low mortality 
risk for their children, even with very short spacing is associated, this effect should
diminish with education level.

Third, having and rearing children involves both direct and time costs, and economic 
factors can, therefore, affect birth spacing \citep{Hotz1997,Schultz1997}.
I divide the economic factors into the effects of price and income shocks,
the expected income trajectory of the husband and wife,
and economies of scale considerations.

The expected effect of economies of scale in childrearing, which can arise, for example, because
childbearing and rearing are time-intensive, but taking care of two young children does not take 
twice as much time as taking care of one child, is shorter intervals between births
\citep{Vijverberg1982,Espenshade1984}.
To the extent that children require the mother to reduce her participation 
in productive activities, parents can lower the cost of children by shortening birth spacing, 
thereby reducing the amount of time the mother is away from productive activities.
In this case, women with higher education will both have lower fertility and shorter
spacing than women with lower education because of the higher opportunity cost of time 
\citep{Ross1974,Newman1981,Newman1984}.
If the mother mainly works at home, there is less incentive to space children for this 
reason and, similarly, higher household income, holding the mother's income constant, 
reduces the incentive to space births for this reason.

@@

The income profile argument assumes that parents know their future income, but income and price shocks may also affect spacing. A temporary, higher cost of women's time
and other inputs into children together with lower income temporarily decreasing fertility 
and, thereby, increase the spacing between births \citep{Moffitt1984,Hotz1988,Portner2001,Alam2018}.
Increasing education likely reduces the risk of income shocks, but the effect on other types of shocks that might impact the spacing is unclear.

Finally, the discussion has so far assumed that children are identical---except for any 
potential differences that arise from spacing---but there is substantial 
evidence that the sex composition of previous children affects birth spacing, with the
absence of boys significantly decreasing the duration to next birth
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.%
\footnote{
There is also evidence of differential stopping behavior, with an additional child more 
likely the fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.
}

It is not, however, immediately apparent that we should expect shorter spacing in the absence 
of boys, if, as discussed above, short spacing lowers the survival probability of the next child.
There are two possibilities for why parents might still try to conceive sooner in the 
absence of boys.%
\footnote{
The fact that there are significant differences in spacing by sex composition indicates that
parents have---at least some---control over the timing of birth.
Hence, it is unlikely that the variations in spacing come about by random chance.
In Matlab, the effect of son preference on birth spacing is stronger in areas with better 
access to family planning \citep{Rahman1993}.
}
One possibility is that parents do not realize that short spacing is potentially
detrimental to the next child's health, resulting in high mortality for both boys and girls 
born after short spacing across all households whether or not they have a son preference.%
\footnote{
We may even see a higher likelihood of dying for sons than girls, given that boys 
have naturally higher mortality.
}

The other possibility is that parents realize that short spacing has adverse health effects, 
but their son-preference is so strong that they are willing to divert substantial resources 
to the next child's health if it is a boy \emph{and} that these investments can 
compensate for the adverse effects.
In this scenario, we should see substantial differences in mortality across sex for short spacing 
children in the absence of any other boys.
If a son is born next, his mortality risk should be substantially lower than expected for 
the spacing duration, and, because this argument relies on redistribution of resources, we should 
expect to see worse outcomes for all the other children (all girls) in the family.
If a girl is born next, there are two possible outcomes.
If parents treat all children/girls the same, we should see a slight worsening in health 
outcomes because fewer resources are available per child; 
if they do not, the last girl will have a much lower likelihood of surviving because the 
parents' incentive is to invest in their prior children rather than the new child.

@@

The higher the cost of prenatal sex determination and abortion,
the less use we should see.
Similarly, any potential adverse effects on the mother and 
offspring from repeated
abortions would also tend to reduce the use of sex selection---
provided that parents are aware of these costs.
Finally, the more abortions between births, the less the opportunity there is to take
advantage of any economies of scale in childrearing.

@@

Given the increase in birth spacing with each abortion, parents may try to conceive even sooner to compensate for the potentially longer spacing and, thereby, still take advantage of economies of scale in childrearing.
By conceiving sooner, they are more likely to be able to take advantage of economies of
scale in childrearing, even if they have to go through one or more abortions.
Furthermore, if health outcomes for short spacing births has improved over time---as is 
likely---we should see an even stronger push toward shorter spacing for women using sex 
selection.

>>>>

This section provides a conceptual framework for understanding what determines  
birth spacing.
I first discuss the main determinants of spacing examined in the prior research, and 
then turn to how son preference and sex selection fits into the decisions on spacing.
Since sex selection and education are closely linked, I pay special attention to how 
education affects spacing.

In the simplest possible dynamic model of fertility, the spacing decision is trivial.
Assume that parents derive benefits from the presence of children, rather than from 
the births themselves, that spacing of children by itself does impact the parental utility, 
and that the parents perfectly control when to have children. 
The optimal decision is then to have the children as early as possible 
\citep{Newman1988}.
The simplest model does not fit the observed spacing patterns, which suggests 
that one or more of the assumptions are wrong.

First, parents do not entirely control the timing of births, but rather decide on 
the levels of sexual activity and contraceptive effort, which, in turn, affect the 
likelihood of conceiving, while taking into account fecundity and their knowledge 
of and access to contraceptive technology.

Fecundity determinants, such as age and health of the mother, are, however, unlikely 
to be significant factors in how spacing changed over time or across education levels 
in India for the period covered.
Outside of direct famine conditions, women's health and nutritional status do not 
appear to affect spacing substantially \citep{Huffman1987,John1987,lindstrom99}.
The Maharashtra drought occurred in 1970--1973 and would, therefore, only impact the first
year or two of the data.
Furthermore, I only examine spacing up to parity fourth, and the age of the mother is,
therefore, unlikely to be a concern when it comes to fecundity, even though 
better-educated women are, on average, older when they marry.

Despite a general belief that better access to and knowledge of contraception 
should lead to longer average spacing, the empirical relationship between contraception 
use and birth spacing is ambiguous 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Yeakey2009,Kim2010,Soest2018}.
This lack of evidence is consistent with the ambiguous theoretical effect of introducing 
modern contraceptives on birth spacing.
On the one hand, in the absence of high-efficacy contraceptive technologies, parents may try to 
avoid overshooting their desired number of children by spacing their children further apart
than they would otherwise do.
Increased reliability of the access to and effectiveness of contraceptives then lead to 
shorter birth intervals because parents are more confident they can effectively
stop childbearing when they want \citep{Keyfitz1971,Heckman1976}.
On the other hand, when contraceptives are ineffective, some birth intervals will
be shorter than ideal, in the way discussed below, through simple random variation.
With increased reliability of contraceptive efforts, parents can better avoid too short 
birth spacing, and average spacing should go up.

To further complicate the relationship between contraception and spacing, the ability 
to successfully use ``low efficacy'' contraceptive methods, such as the rhythm method, 
is increasing in education, while there is no difference across education levels 
when it comes to modern contraceptives such as pills or injectables \citep{Rosenzweig1989}.
Additionally, even in the absence of modern contraceptives, families appear able to 
time or at least postpone births when they have substantial enough incentive 
\citep{Jayachandran2011,Alam2018}.

Second, parents may care about the spacing between birth because of the 
potential effects of spacing on the children.%
\footnote{
In principle, the mother could also be affected, but the estimated impact on 
mothers' anthropometric status is mixed, and there appears to be no 
effect on mortality 
\citep{Ronsmans1998,Menken2003,Dewey2007,Conde-Agudelo2012}.
}
One theory is that child ``quality'' increases with spacing 
since parental attention per child increases
\citep{Zajonc1975,Zajonc1976,Razin1980}.
In this case, birth spacing should increase in family income, while the 
effect of maternal education is ambiguous since higher education, on the 
one hand, makes the mother's time spent out of the labor market more 
costly and on the other hand increases how productive the mother is 
in childrearing.
The evidence on spacing's effect on child quality measures such as IQ 
and education is, however, mixed for developed countries and non-existing 
for developing countries
\citep{Powell1993,Pettersson-Lidbom2009,Buckles2012,Barclay2017}.

For one crucial aspect of child quality---health and mortality---shorter spacing 
does lead to worse outcomes, especially for very short intervals of 24 months 
or less, although this effect decreases with maternal education
\citep{Whitworth2002,Conde-Agudelo2006,Conde-Agudelo2012,Molitoris2019}.%
\footnote{
The underlying mortality risk could also affect birth spacing, but
there is little empirical evidence for this 
\citep{Newman1983,Newman1988,Bhalotra2008}.
}
If parents realize short birth spacing increases their children's mortality risk, 
they likely prefer to space births further apart.
However, this effect should diminish with education level since women with 
secondary or tertiary education experience relatively low mortality risk for 
their children, even with very short spacing is associated.

Third, parents incur both direct and time costs when they have children, 
and economic factors can, therefore, affect birth spacing \citep{Hotz1997,Schultz1997}.
I divide the economic factors into the effects of price and income shocks,
the expected income trajectory of the husband and wife,
and economies of scale considerations.

The expected effect of economies of scale in childrearing, which, for example, 
can arise because taking care of two young children does not take twice as much 
time as taking care of one child, is to shorten intervals between births
\citep{Vijverberg1982,Espenshade1984}.
To the extent that children require the mother to do less market work, parents 
can lower the cost of children by reducing birth spacing.
In this case, better-educated women will both have lower fertility and shorter
spacing than less-educated women because of their higher opportunity cost of time 
\citep{Ross1974,Newman1981,Newman1984}.
However, if the mother mainly works at home or as household income increases, there 
is less incentive to space children for this reason.

@@

The income profile argument assumes that parents know their future income, 
but income and price shocks may also affect spacing. 
A temporary, higher cost of women's time or a lower household income, will 
temporarily decrease fertility and, thereby, increase birth spacing 
\citep{Moffitt1984,Hotz1988,Portner2001,Alam2018}.
Increasing education likely reduces the risk of income shocks, but the 
effect on other types of shocks that might impact the spacing is unclear.

Finally, the discussion has so far assumed that children are identical---except 
for any differences that arise from spacing---but there is substantial evidence 
that the sex composition of previous children affects birth spacing, with the
absence of boys significantly decreasing the duration to next birth
\citep{Haughton1995,Haughton1996,Rahman1993,Bhalotra2008,Kumar2016,Soest2018}.
There is also evidence of differential stopping behavior, with an additional child 
more likely the fewer sons a family has 
\citep{repetto72,Das1987,Arnold1997,arnold98,clark00,Basu2010,Barcellos2014}.

It is not, however, immediately apparent that we should expect shorter spacing in 
the absence of boys, if short spacing lowers the survival probability of the 
next child, as discussed above.
There are two potential reasons why parents might still try to conceive sooner 
in the absence of boys.%
\footnote{
The significant differences in spacing by sex composition indicate that
parents control the timing of birth, at least to some extent.
In Matlab, the effect of son preference on birth spacing is stronger in 
areas with better access to family planning \citep{Rahman1993}.
}
One possibility is that parents do not realize that short spacing is potentially
detrimental to the next child's health, resulting in high mortality for both boys 
and girls born after short spacing across all households whether or not they have 
a son preference.%
\footnote{
We may even see a higher likelihood of dying for sons than girls, given that boys 
have naturally higher mortality.
}

The other possibility is that parents realize that short spacing has adverse 
health effects, but their son-preference is so strong that they are willing to 
divert substantial resources to the next child if it is a boy.
Assuming that these investments can compensate for the adverse effects,
we should see significant differences in mortality across sex for short-spacing 
children in the absence of boys.
If a son is born, his mortality risk should be substantially lower than expected 
for the spacing duration, and we should expect worse outcomes for the other 
children (all girls).
If a girl is born, there are two possible outcomes.
If parents treat all children/girls the same, we should see a slight worsening in 
health outcomes because fewer resources are available per child; 
if they do not, the last girl will have a much lower likelihood of surviving 
because the parents' incentive is to invest in their prior children rather than 
the new child.

@@

The higher the cost of prenatal sex determination and abortion,
the less use we should see.
Similarly, any potential adverse effects on the mother and offspring 
from repeated abortions would also tend to reduce the use of sex 
selection---provided that parents are aware of these costs.
Finally, the more abortions between births, the less the opportunity 
there is to take advantage of any economies of scale in childrearing.

@@

Given the increase in birth spacing with each abortion, parents may try to 
conceive even sooner to compensate for the potentially longer spacing and, 
thereby, still take advantage of economies of scale in childrearing.
Furthermore, if health outcomes for short spacing births have improved over 
time---as is likely---we should see an even stronger push toward shorter 
spacing for women using sex selection.
